---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# typed

Experiment on static typing in R. Largely untested!

We use the question mark operator to support static typing in R.

3 features are proposed :

We use it for 3 different things here :

* set variable types in a script or the body of a function
* set argument typesin a function definition
* set return type of a function

We detail those below.


## Installation

Install with:

``` r
remotes::install_github("moodymudskipper/typed")
```

And attach with : 

```{r}
# masking warning about overriding `?`
library(typed, warn.conflicts = FALSE) 
```

## set variable type

We can set a variable type explicitly or implicitly. The following are equivalent.

```{r, eval = FALSE}
# implicit
? x <- 1
```

```{r, eval = FALSE}
# explicit
numeric ? x <- 1
```

```{r, eval = FALSE}
# explicit without assignment on declaration
numeric ? x
x <- 1
```

Let's test it

```{r}
numeric ? x
x
x <- 1
x
x <- "a"
```

To assess if assignments are allowed we compare `vctrs::vec_ptype(assigne_value)`
to the prototype given on the left hand side, which we can give as a function
or value. in the latter for instance `vctrs::vec_ptype(1)` and `numeric()` return
the same thing so the assignment is allowed.

Using *{vctrs}* allows us more flexibility than working with `class()` or `type_of()`.

```{r}
library(vctrs)
new_date ? my_date <- "2020-10-16"
new_date ? my_date <- as.Date("2020-10-16")

val <- new_list_of(list(1, 2), numeric())
val
new_list_of(list(), numeric()) ? new_val <- list(1, 2)
new_list_of(list(), numeric()) ? new_val <- val
```

We can also define constants:

```{r}
character ? chr1 <- const(1) 
character ? chr1 <- const("a") 
chr1
chr1 <- "b"
```

## set argument type

We can set argument types this way : 


```{r}
? add <- function (x = ?numeric, y = 1 ?numeric) {
  x + y
}
```

Note that we started the definition with a `?`, and that we gave a default to
`y`, but not `x`

This created the following function :

```{r}
add
```

You can see a call to `assert_types()`, it ensures arguments passed have
the right type, and that they won't be overridden by a different type.

Let's test it.

```{r}
add(2, 3)
add(2)
add("a")
```

Let's create a function that tried to modify `y` and assign the wrong type

```{r}
? add_wrong <- function (x = ?numeric, y = 1 ?numeric) {
  y <- as.character(y)
  x + y
}
add_wrong(2)
```

## set function return type

Setting a function return type is similar to setting variable types, but the function
should only return an object named `res`, which cannot ever have another type than
the one we set.

Putting it all together we have :

```{r}
numeric ? add10 <- function (x = ?numeric) {
  ? y <- 10
  res <- x + y
  res
}
add10
add10(20)
```


## Notes

* This is inspired in good part by Jim Hester and Gabor Csardi's work.
* The magic comes from the fact that apparent variables are made into active
  bindings, using a variation of the last example of `?bindenv`
* Your package would import *{typed}* but `?` won't be exposed to the user,
they will see it in the code but will be able to use `?` just as before.
* It might be slow
* We use `vctrs::vec_ptype` on the given prototype and the object to ensure their
  compatible, it's not always satisfactory. expressions or functions ar not
  supported, the prototype of a factor necessarily contains the levels, the prototype
  of a dataframe necessarily contains the column names. I'd like to be more flexible (suggestions welcome).
* In particular, setting length as part of the type would be great, so we could
  define scalars for instance.
