---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# typed

Experiment on static typing in R. Not robust!

I realize the syntax is not quite right, but publishing a draft already.
Output type of the function is not enforced yet, only arguments and variables
defined in the function for now.

## Installation

Install with:

``` r
remotes::install_github("moodymudskipper/typed")
```

## Example

We define a function below, the suffix after `..` describes the expected
prototype of the output or arguments. These can be any of "list", "logical", 
"integer", "numeric", "double", "complex", "character" "raw", with the addition
of default prototypes proposed by *{vctrs}*, such as "date", "datetime" etc.

`:=` must be used to define the function.
Inside the body `y := 1` makes sure `y` returns 1, and that it can only be assigned
doubles (the prototype of `1`).

```{r}
library(typed)
add1..numeric := function (x..numeric) {
  y := 1
  x + y
}
```

This creates a function `add1()`, with attributes and a call `assert_types()`.

```{r}
add1
```

The call to assert_types will check that all parameters are right, and will
make sure they can't be overriden by a value having another prototype.

We can call it :

```{r}
add1(3)
```

Fails early if fed the wrong arg type :

```{r}
add1("a")
```

if we had changed the value of x, by assigning a char in the body it would
have failed too, right when we'd try to assign it:

```{r}
add1..numeric := function (x..numeric) {
  message("we get up to here")
  x <- as.character(x)
  message("but not there")
  y := 1
  x + y
}

add1(3)
```


Same thing if we try to modify `y` using an illegal type:

```{r}
add1..numeric := function (x..numeric) {
  y := 1 # this forces y keep the same prototype
  message("we get up to here")
  y <- "2"
  message("but not there")
  x + y
}

add1(3)
```
