---
output: github_document
---

<!-- badges: start -->
<!--[![Travis build status](https://travis-ci.com/moodymudskipper/typed.svg?branch=iteration2)](https://travis-ci.com/moodymudskipper/typed)
<!-- badges: end -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# typed

*{typed}* implements static typing in R, it has 3 main features:

* set variable types in a script or the body of a function
* set argument types in a function definition
* set return type of a function

The user can define their own types, or leverage assertions from
other packages.

Under the hood we use active bindings, so once a variable is restricted to
a type, or by any other condition, it cannot be modified in a way that would
not respect these restrictions.

## Installation

Install with:

``` r
remotes::install_github("moodymudskipper/typed@iteration2")
```

And attach with : 

```{r}
# masking warning about overriding `?`
library(typed, warn.conflicts = FALSE) 
```

## Set variable type

### Question mark notation and `declare`

Here are examples on how we would set types 

```{r}
Character() ? x # restrict x to "character" type
x <- "a"
x

Integer(3) ? y <- 1:3 # restrict y to "integer" type of length 3
y
```

We cannot assign values of the wrong type to `x` and `y` anymore.

```{r}
x <- 2

y <- 4:5

y[1] <- 10
```

But the right type will work.

```{r}
x <- c("b", "c")

y <- c(1L, 10L, 100L)
```

If you dislike the `?` operator you can use `declare`, it's a strict equivalent,
slightly more efficient, which looks like `base::assign`.

```{r}
declare("x", Character())
x <- "a"
x

declare("y", Integer(3), 1:3)
y
```

### assertion factories and assertions

`Integer` and `Character` are function factories (functions that return functions),
thus `Integer(3)` and `Character()` are functions.

The latter functions operate checks on a value and in case of success 
return this value, generally unmodified. For instance :

```{r}
Integer(3)(1:2)

Character()(3)
```

We call `Integer(3)` and `Character()` assertions, and we call
`Integer` and `Character` assertion factories.

The package contains many assertion factories (see `?native_types`), 
the main ones are: 

* `Any` (No default restriction)
* `Logical`
* `Integer`
* `Double`
* `Character`
* `List`
* `Environment`
* `Factor`
* `Matrix`
* `Data.frame`
* `Date`
* `Time` (POSIXct)

### Custom types

As we've seen with `Integer(3)`, `Data.frame(ncol = 5)`, passing arguments to a
assertion factory restricts the type. For instance `Integer` has arguments `length` and
`...`, in the dots we can use arguments named as functions and with the value of
the expected result.

```{r}
Integer(anyNA = FALSE) ? x <- c(1L, 2L, NA)
```

Useful arguments might be for instance, `anyDuplicated = 0L`, `names = NULL`, 
`attributes = NULL`... Any available function can be used.

That makes assertion factories very flexible! If it is still not flexible enough, one
can provide conditions using formulas.

```{r}
fruit <- Character(1, "`value` is not a fruit!" ~ . %in% c("apple", "pear", "cherry"))

fruit ? x <- "potatoe"
```

### Leverage assertions from other packages, build your own assertion factories

Some great packages provide assertions, and they can be used with `typed`
provided that they take the object as a first input and return the object
if no failure. Richie Cotton's *{assertive}* and Michel Lang's *{checkmate}* 
both qualify and we'll use them as examples.

```{r}
library(assertive)
assert_is_monotonic_increasing ? z
z <- 3:1
```

If we want to use more than the first argument, we should create an assertion factory :

```{r}
Monotonic_incr <- as_assertion_factory(assert_is_monotonic_increasing)
Monotonic_incr(strictly = TRUE) ? z
z <- c(1, 1, 2)
library(checkmate)
```

`as_assertion_factory` can be used to create your own assertion factories,
in fact
[it's used to build the native assertion factories of this package](https://github.com/moodymudskipper/typed/blob/iteration2/R/06_native_types.R)
.

### Constants

To define a constant, we just surround the variable by parentheses (think of 
them as a protection)

```{r}
Double() ? (x) <- 1
x <- 2

? (y) <- 1
y <- 2
```

## Set argument type

We can set argument types this way : 


```{r}
add <- ? function (x= ? Double(), y= 1 ? Double()) {
  x + y
}
```

Note that we started the definition with a `?`, and that we gave a default to
`y`, but not `x`

This created the following function, by adding declarations at the top of the
function.

```{r}
add
```

Note that a declaration will fail if the variable already exists in the
environment, but an exception is made for unevaluated function arguments.

Let's test it by assigning a right and wrong type

```{r}
add(2, 3)
add(2, 3L)
```

Arguments `x` and `y` will also be restricted to the type "integer" in the rest of
the body of the function.

## Set function return type

To set a return type one way is to manually use the `type_checker() ? return(value)`
syntax with every return value, as we do here (note that we need the `?` or the
precedence of `?` will cause issues) :

```{r}
add_or_substract <- function (x, y, substract = FALSE) {
  if(substract) {Double() ? return(x - y)}
  Double() ? return(x + y)
}
add_or_substract(1, 2)
add_or_substract(1L, 2L)
```

A simpler way is to use `?` before the function definition as in the previous section, 
but to give it a assertion factory on the left hand side. It will ultimately create
the same function code, but the source code is more readable to read, we don't
need to worry about precedence, and the function will get a class "type" and attributes.

```{r}
add_or_substract <- Double() ? function (x, y, substract = FALSE) {
  if(substract) return(x - y)
  x + y
}
add_or_substract
```

These can be defined in a package and documented with *{roxygen2}* like regular
functions, except that you'll need to make sure to add the `@name` tag.

```{r, eval = FALSE}
#' add
#'
#' @param x double
#' @param y double
#' @export
#' @name add
add <- Double() ? function (x = ? Double(), y = 1 ? Double()) {
  x + y
}
```


## Notes

* This is inspired in good part by Jim Hester and Gabor Csardi's work.
* The magic comes from the fact that apparent variables are made into active
  bindings, using a variation of the last example of `?bindenv`
* Your package would import *{typed}* but `?` won't be exposed to the user,
they will see it in the code but will be able to use `?` just as before. In fact
the most common standard use `?mean` still works even when *{typed}* is attached.
