---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# typed

Experiment on static typing in R. Largely untested!

## Installation

Install with:

``` r
remotes::install_github("moodymudskipper/typed")
```

## Example

We define a function below, the question mark is used to introduce static typing,
`numeric` here is a prototyping function as understood by *{vctrs}*, 
(we could have written `numeric()` with the same effect, and we could use *{vctrs}*
functions such as `new_date()` too).

We use it for 3 different things here :

* set a return type to the function `add2` : if the first `?` has a lhs, the function
  should only return a variable `res`, and *{typed}* will make sure that `res`
  is always of the right type.
* set argument type : the arguments for which we want to set a type are suffixed
  with `?`, in the case below we don't have a default value so `?` has no lhs
* set variable type : we initiate the variable `y` as numeric in this example
  and assigning any non numeric to it will fail.

```{r}
library(typed)
numeric ? add2 <- function (x = ?numeric) {
  numeric ? y
  y <- 2
  res <- x + y
  res
}
```

This creates a function `add2()`, with attributes and a call `assert_types()` at
the start of the body.

```{r}
add2
```

The call `assert_types()` will check that all parameters are right, and will
make sure they can't be overridden by a value having another prototype.

We can call it :

```{r}
add2(3)
```

It fails early if fed the wrong arg type :

```{r}
add2("a")
```

if we had changed the value of x, by assigning a char in the body it would
have failed too, right when we'd try to assign it:

```{r}
numeric ? add2 <- function (x = ?numeric) {
  message("we get up to here")
  x <- as.character(x)
  message("but not there")
  numeric ? y
  y <- 2
  res <- x + y
  res
}

add2(3)
```


Same thing if we try to modify `y` using an illegal type:

```{r}
numeric ? add2 <- function (x = ?numeric) {
  numeric ? y
  y <- 2
  message("we get up to here")
  y <- "2"
  message("but not there")
  res <- x + y
  res
}

add2(3)
```

instead of : 

```{r, eval = FALSE}
  numeric ? y
  y <- 2
```

We can write on one line : 

```{r, eval = FALSE}
  numeric ? y <- 2
```

It will test the value of y just like with the former syntax.

We can also set the prototype implicitly by simply typing : 

```{r, eval = FALSE}
  ? y <- 2
```


Let's write a variant, we might decide not to set a return type, 
not to declare our y variable, and just set argument types, setting a default
this time.


```{r}
? add2 <- function (x = 1 ?numeric) {
  y <- 2
  x + y
}

add2()
```

It fails if we try to override the argument with a different type:

```{r}
? add2 <- function (x = 1 ?numeric) {
  y <- 2
  x <- as.character(x)
  x + y
}

add2()
```

## Notes

* This is inspired in good part by Jim Hester and Gabor Csardi's work.
* The magic comes from the fact that apparent variables are made into active
  bindings, using a variation of the last example of `?bindenv`
* Your package would import *{typed}* but `?` won't be exposed to the user,
they will see it in the code but will be able to use `?` just as before.
* It might be slow
* We use `vctrs::vec_ptype` on the given prototype and the object to ensure their
  compatible, it's not always satisfactory. expressions or functions ar not
  supported, the prototype of a factor necessarily contains the levels, the prototype
  of a dataframe necessarily contains the column names. I'd like to be more flexible.
* In particular, setting length as part of the type would be great, so we could
  define scalars for instance.
* A class and printing method for these functions would be nice, to print the
attributes in a prettier way, and to print the `?` better, maybe some fancy
*{crayon}* stuff to highlight types.
